---
title: " Study II - Beta-diversity analyses of metaphlan profiles"
author: "Florentin Constancias"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output: 
  html_document: 
    toc: yes
    keep_md: yes
---


```{r setup, include=FALSE}
rm(list = ls())
gc()

# invisible(lapply(paste0('package:', names(sessionInfo()$otherPkgs)), detach, character.only=TRUE, unload=TRUE))

# options(java.parameters = "-Xmx80000m")
```


```{r setup2, include=FALSE}
knitr::opts_chunk$set(warning = FALSE)
knitr::opts_chunk$set(message = FALSE)
##knitr::opts_chunk$set(cache=TRUE)
# knitr::opts_chunk$set(echo = TRUE)
# knitr::opts_chunk$set(dev = "ragg_png") # https://ragg.r-lib.org/index.html
##knitr::opts_chunk$set(fig.width = 6) #https://r4ds.had.co.nz/graphics-for-communication.html#figure-sizing
knitr::opts_chunk$set(fig.asp = 0.618)
knitr::opts_chunk$set(fig.show = "hold")
knitr::opts_chunk$set(fig.show = "70%")
knitr::opts_chunk$set(fig.align = "center")
```

```{r setup3, include=FALSE}
library(tidyverse)
library(readxl)
library(readr)
library(speedyseq)
library(microViz)
```

```{r setup4, include=FALSE}

'%!in%' <- function(x,y)!('%in%'(x,y))

source_dir = "https://raw.githubusercontent.com/fconstancias/DivComAnalyses/master/R/"
# source_dir = "~/Documents/GitHub/DivComAnalyses/R/"

source(paste0(source_dir,"phyloseq_normalisation.R"))
source(paste0(source_dir,"phyloseq_varia.R"))
source(paste0(source_dir,"phyloseq_beta.R"))
```


```{r}
load(here::here("../../data/processed_data/metaphlan/01_data.Rdata"))
load(here::here("../../data/processed_data/metaphlan/02_data.Rdata"))
```

```{r}
pd <- position_dodge(0.3)
```

Let's source the function:

```{r, include=TRUE}
source("https://raw.githubusercontent.com/fconstancias/KU_Caroline/refs/heads/main/code/functions/phyloseq_functions.R")

# details:

#' @title This function calculates beta diversity, generates ordination plots (e.g., PCoA),  performs PERMANOVA, and optionally creates boxplots of distances between metadata groups.
#' @author Florentin Constancias
#' @param ps_up A phyloseq object containing microbiome data.
#' @param beta A list of distance matrices for beta diversity (e.g., Bray-Curtis, Aitchison).
#' @param m Character string specifying the ordination method (default: "PCoA").
#' @param color_group Character string for the variable used to color points in ordination plots (default: "Time").
#' @param shape_group Character string for the variable used to shape points in ordination plots (default: "Sample").
#' @param alpha Optional variable for transparency adjustment in ordination plots (default: NULL).
#' @param col_pal Color palette for point colors (default: `time_pal`).
#' @param fill_pal Fill color palette for point fills (default: `time_pal`).
#' @param path_group Character string for the variable grouping paths in ordination plots (default: "interaction(Sample,Subject)").
#' @param facet_formula Formula for faceting ordination plots (default: "Sample ~ .").
#' @param axis1 Integer specifying the ordination axis to plot on the x-axis (default: 1).
#' @param axis2 Integer specifying the ordination axis to plot on the y-axis (default: 2).
#' @param seed Integer for random seed, ensuring reproducible results in ordination (default: 123).
#' @param permanova_terms Character vector of variables for PERMANOVA tests (default: c("Time", "cluster_Dtp2")).
#' @param metadata_dist_boxplot Optional variable for generating distance boxplots (default: NULL).
#' @param strata Variable for stratified PERMANOVA testing (default: "none").
#' @param perm Number of permutations for PERMANOVA (default: 999).
#' 
#' @return A list of results, including:
#'   - `pcoas`: Ordination plots with faceting and custom color/shape groups.
#'   - `expl_var`: Explained variance for each axis in ordination plots.
#'   - `PCOA`: Detailed ordination plot for the rAitchison distance with paths.
#'   - `PCOA_leg`: Legend for the ordination plot.
#'   - `perm`: PERMANOVA results.
#'   - `pw_perm`: Pairwise PERMANOVA results for each variable in `permanova_terms`.
#'   - `tw_perm`: Two-way PERMANOVA results for each variable in `permanova_terms`.
#'   - Optional: `dist_boxplot` if `metadata_dist_boxplot` is provided, showing boxplots of distances.

```

# Compute the distance for all samples:

```{r, warning=FALSE, echo=FALSE}
ps_up %>%
  subset_taxa(Class != "UNCLASSIFIED") %>% 
  transform_sample_counts(function(x) x/sum(x) * 1) %>%
  phyloseq_compute_bdiv() -> beta


ps_up %>% 
  subset_taxa(Class != "UNCLASSIFIED") %>% 
  transform_sample_counts(function(x) x/sum(x) * 1) %>%
  microViz::dist_calc(., dist = "robust.aitchison") %>% 
  microViz::dist_get() %>% 
  magrittr::divide_by(100) -> beta$rAitchison

beta$bray <- NULL
beta$sorensen <- NULL
```


# Run fonction for full dataset:

```{r, warning=FALSE, echo=FALSE}
ps_up %>%
  subset_taxa(Class != "UNCLASSIFIED") %>% 
  transform_sample_counts(function(x) x/sum(x) * 1) %>% 
  phyloseq_explore_beta(ps_up = .,
                        beta = beta,
                        color_group = "Time",
                        shape_group = "Sample",
                        distance_for_more = "rAitchison",
                        metadata_sel = c("delta_mean_plaque", "delta_mean_bleeding"),
                        pval_cutoff = 0.05,
                        top_r = 12,
                        alpha = NULL,
                        col_pal = time_pal,
                        fill_pal = time_pal,
                        path_group = "interaction(Sample,Subject)",
                        facet_formula = "Sample ~ .",
                        axis1 = 1,
                        axis2 = 2,
                        seed = 123,
                        perm = 999,
                        permanova_terms = c("Sample", "Time", "cluster_Dtp2"),
                        metadata_dist_boxplot = c("Subject", "Time", "Sample"),
                        strata = "none") -> out
```


Explore PCoA : distances show similar patterns

```{r}
out$pcoas
```

Explained variance of the plots above.

```{r}
out$expl_var %>% 
  DT::datatable()
```

Focus on Robust Aitchinson distance, clear discrimination between site, display trajectories.

```{r}
out$PCOA$p + facet_null() + stat_ellipse(data=out$PCOA$p$data,
                                         segments = 21, linetype = 2, color = "grey80",
                                         aes(group = interaction(time, Sample)),
)
```
Corresponding legend:

```{r}
out$PCOA$p_leg
```


```{r}
out$env_fit_tax$plot 
```

```{r}
out$env_fit_meta$plot
```

Statistical evaluation of the beta diversity: association with Sample_Type, Time and Cluster of inflamation at TP2 -PERMANOVA:

Sample Type explain most of the variation (R2) > Time > Cluster

More info: https://gist.github.com/claczny/3415270a6c6919969bff79d2c246a527

```{r}
out$perm %>% 
  mutate_if(is.double, ~round(., 3)) %>% 
  dplyr::filter(Distance == "rAitchison", terms_margin == "terms") %>% 
  DT::datatable()
```

Statistical evaluation of the beta diversity - Pairwise PERMANOVA tests between Time, Sample Type, Inflamation response clusters:


```{r}
out$pw_perm %>% 
  dplyr::filter(Distance == "rAitchison") %>% 
  mutate_if(is.numeric, ~round(., 3)) %>% 
  DT::datatable()
```

We could also play the intra-subject distance accross time | maybe with within group variation for TP1: within TP1 / Sample:

```{r}
out$dist_box
```


Statistical evaluation of the beta diversity - Pairwise Multivariate T.tests between Time, Sample Type, Inflamation response clusters:

```{r}
# out$tw_perm %>% 
#   DT::datatable()
```

We can also explore the groups of the different variables:

Sample

```{r}
out$Sample$Saliva
```

```{r}
out$Sample$Plaque
```
```{r}
out$Time$TP1
```

```{r}
out$Time$TP2
```

```{r}
out$Time$TP3
```


# Now we can explore more Saliva and Plaque samples:


## Saliva:

```{r,return = FALSE}
ps_up %>%
  subset_samples(Sample == "Saliva") %>% 
  subset_taxa(Class != "UNCLASSIFIED") %>% 
  transform_sample_counts(function(x) x/sum(x) * 1) %>% 
  microViz::ps_mutate(cluster_Dtp2 = as.factor(cluster_Dtp2)) %>% 
  compute_plot_beta(ps_up = .,
                    beta = beta,
                    color_group = "Time",
                    shape_group = "Sample",
                    alpha = NULL,
                    col_pal = time_pal,
                    fill_pal = time_pal,
                    path_group = "interaction(Subject)",
                    facet_formula = "Sample ~ .",
                    axis1 = 1,
                    axis2 = 2,
                    seed = 123,
                    permanova_terms = c("Time","cluster_Dtp2"),
                    # metadata_dist_boxplot = NULL ,
                    strata = "none") -> saliva
```


```{r}
saliva$pcoas
```


```{r}
saliva$PCOA
```

```{r}
saliva$perm %>% 
  mutate_if(is.double, ~round(., 3)) %>% 
  dplyr::filter(Distance == "rAitchison", terms_margin == "terms") %>%
  DT::datatable()
```


Pairwise comparisons between groups: Time, ... 


Interesting to see that when considering only presence and absence of taxa (i.e., binary bJaccard), no significant differences between TP1 and TP3, but still when considering proportion of taxa. 


```{r}
saliva$pw_perm %>% 
  # dplyr::filter(Distance == "rAitchison") %>% 
  mutate_if(is.numeric, ~round(., 3)) %>% 
  DT::datatable()
```

```{r}
# saliva$tw_perm %>% 
#   DT::datatable()
```

We could test if the distance between saliva and plaque per individual changes over time:

We have this distance to the baseline boxplot - which is the same as prev. but only Saliva here. But for this we need to use the ouputs generated with the full dataset.

```{r}
saliva$dist_box
```

During perturbation : TP2 : saliva and Plaque microbiome are similar:

```{r}
out$dist_df %>% 
  dplyr::filter(Sample_1 != Sample_2,
                Time_1 == Time_2,
                Subject_1 == Subject_2) %>%
  arrange(Distance, Subject_2, Time_2) %>% 
  ggplot(data = ., aes_string(x="Time_2", y="value")) +
  geom_boxplot(outlier.colour = NA, alpha=0.7, aes_string(fill = "Time_2")) +
  geom_jitter(size=1, position = pd, aes_string(color = "Subject_1")) +
  geom_line(aes_string(group="Subject_1"), position = pd, linetype = "dashed", color = "grey50", linewidth = 0.08) +
  facet_grid(as.formula(paste0("Distance ~ Sample_1")), scales = "free_y", space = "fixed", switch = "y") +
  scale_color_manual(name = "", values = sub_pal,
                     na.value = "black") +
  scale_fill_manual(name = "", values = time_pal,
                    na.value = "black") +
  theme_light() + ylab("Between Site Intra-individual distance ") + xlab(NULL) + theme(
    axis.text.x = element_blank()) +  theme_linedraw() + theme(strip.placement = "outside") -> d_box

d_box %>%
  ggpubr::get_legend(.) %>%
  ggpubr::as_ggplot(.) -> d_box_leg

d_box + theme(legend.position = "none") -> d_box

d_box
```

## Plaque:

```{r,return = FALSE, message=FALSE, warning=FALSE}
ps_up %>%
  subset_samples(Sample == "Plaque") %>% 
  subset_taxa(Class != "UNCLASSIFIED") %>% 
  transform_sample_counts(function(x) x/sum(x) * 1) %>% 
  microViz::ps_mutate(cluster_Dtp2 = as.factor(cluster_Dtp2)) %>% 
  compute_plot_beta(ps_up = .,
                    beta = beta,
                    color_group = "Time",
                    shape_group = "Sample",
                    alpha = NULL,
                    col_pal = time_pal,
                    fill_pal = time_pal,
                    path_group = "interaction(Subject)",
                    facet_formula = "Sample ~ .",
                    axis1 = 1,
                    axis2 = 2,
                    seed = 123,
                    permanova_terms = c("Time","cluster_Dtp2"),
                    # metadata_dist_boxplot = NULL ,
                    strata = "none") -> plaque
```


```{r}
plaque$pcoas
```


```{r}
plaque$PCOA
```

```{r}
plaque$perm %>% 
  mutate_if(is.double, ~round(., 3)) %>% 
  dplyr::filter(Distance == "rAitchison", terms_margin == "terms") %>%
  DT::datatable()
```

Pairwise comparisons between groups: Time, ... 

Interesting to compare between saliva and plaque... Maybe a plot displaying R2 and adjpavalue?


```{r}
plaque$pw_perm %>% 
  # dplyr::filter(Distance == "rAitchison") %>% 
  mutate_if(is.numeric, ~round(., 3)) %>% 
  DT::datatable()
```

```{r}
plaque$tw_perm %>% 
  DT::datatable()
```

Display PERMANOVA results as a plot - in progress:

```{r}
permanova_terms = c("Time","cluster_Dtp2")


saliva$perm %>% 
  mutate(Data = "Saliva") %>% 
  rbind(.,
        plaque$perm %>% 
          mutate(Data = "Plaque")    
  ) %>% 
  filter(!terms %in% c("Residual","Total")) %>%
  mutate(terms = fct_relevel(terms, c(permanova_terms, "Time:cluster_Dtp2"))) %>%
  # arrange(Distance, Group2) %>%
  # filter(p.adj <= 0.05) %>%
  # mutate(p.adj = replace(p.adj, p.adj > 0.05, NA)) %>% 
  ggplot(mapping = aes(x = terms,
                       y = R2, #`Pr(>F)` ,
                       size = R2,
                       # fill = `Pr(>F)`,
                       color = `Pr(>F)`)) + #,
  # size = R2)) +
  facet_grid(Distance + Data ~ terms_margin+ Group,  scales = "fixed", space = "fixed", drop = FALSE) +
  geom_point(alpha=0.8) +
  # geom_tile() +
  # geom_bar(stat="identity") +
  theme_linedraw() + 
  xlab(NULL) + ylab(NULL) +
  scale_size(range = c(1, 8), 
             breaks = c(0.05, 0.1,0.2,0.6),
             labels = c(0.05, 0.1,0.2,0.6),
             limits = c(0,1),
             # trans = "sqrt",
             name="PERMANOVA R2 value") + 
  scale_color_viridis_c(name = "adj p.value",
                        na.value = "grey10",
                        trans = scales::pseudo_log_trans(sigma = 0.001),
                        limits = c(0,0.05),
                        breaks = c(1,0.05, 0.01, 0.001),
                        labels = c(1,0.05, 0.01, 0.001)) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5)) #+
# coord_equal()



```

Site across time?

```{r,return = FALSE, message=FALSE, warning=FALSE}
tmp = ps_up
group = "Time"
Time_out=NULL
# Time_out <- vector("list", length(tmp %>%
#                                get_variable(group) %>%
#                                levels()))
# names(Time_out) <- tmp %>%
#   get_variable(group) %>% levels()

for(tp in tmp %>%
    get_variable(group) %>%
    unique()){
  
  prune_samples(get_variable(tmp, group) == tp,
                tmp) %>%
    subset_taxa(Class != "UNCLASSIFIED") %>% 
    transform_sample_counts(function(x) x/sum(x) * 1) %>% 
    microViz::ps_mutate(cluster_Dtp2 = as.factor(cluster_Dtp2)) %>% 
    compute_plot_beta(ps_up = .,
                      beta = beta,
                      color_group = "Sample",
                      shape_group = "Sample",
                      alpha = NULL,
                      col_pal = sample_pal,
                      fill_pal = sample_pal,
                      path_group = "interaction(Subject)",
                      facet_formula = "Sample ~ .",
                      axis1 = 1,
                      axis2 = 2,
                      seed = 123,
                      permanova_terms = c("Sample"),
                      metadata_dist_boxplot = NULL ,
                      strata = "none") -> Time_out[[tp]]
}

```


```{r}
Time_out %>% 
  plyr::ldply(., function(x) x$perm) %>%
  rename(Time = '.id') %>% 
  DT::datatable()
```


At baseline, TP2, TP3: impact of clusterdtp2 in the diff site?


```{r,return = FALSE, message=FALSE, warning=FALSE}
tmp = ps_up
group = "Time"
Time_out_clust=NULL
# Time_out <- vector("list", length(tmp %>%
#                                get_variable(group) %>%
#                                levels()))
# names(Time_out) <- tmp %>%
#   get_variable(group) %>% levels()

for(tp in tmp %>%
    get_variable(group) %>%
    unique()){
  
  prune_samples(get_variable(tmp, group) == tp,
                tmp) %>%
    subset_taxa(Class != "UNCLASSIFIED") %>% 
    transform_sample_counts(function(x) x/sum(x) * 1) %>% 
    microViz::ps_mutate(cluster_Dtp2 = as.factor(cluster_Dtp2)) %>% 
    compute_plot_beta(ps_up = .,
                      beta = beta,
                      color_group = "Sample",
                      shape_group = "Sample",
                      alpha = NULL,
                      col_pal = sample_pal,
                      fill_pal = sample_pal,
                      path_group = "interaction(Subject)",
                      facet_formula = "Sample ~ .",
                      axis1 = 1,
                      axis2 = 2,
                      seed = 123,
                      permanova_terms = c("Sample", "cluster_Dtp2"),
                      metadata_dist_boxplot = NULL ,
                      strata = "none") -> Time_out_clust[[tp]]
}

```

```{r}
Time_out_clust %>% 
  plyr::ldply(., function(x) x$perm) %>%
  rename(Time = '.id') %>% 
  DT::datatable()
```

Saliva

```{r,return = FALSE, message=FALSE, warning=FALSE}
tmp = ps_up %>% 
  subset_samples(Sample == "Saliva")
group = "Time"
Time_out_saliva_clust=NULL
# Time_out <- vector("list", length(tmp %>%
#                                get_variable(group) %>%
#                                levels()))
# names(Time_out) <- tmp %>%
#   get_variable(group) %>% levels()

for(tp in tmp %>%
    get_variable(group) %>%
    unique()){
  
  prune_samples(get_variable(tmp, group) == tp,
                tmp) %>%
    subset_taxa(Class != "UNCLASSIFIED") %>% 
    transform_sample_counts(function(x) x/sum(x) * 1) %>% 
    microViz::ps_mutate(cluster_Dtp2 = as.factor(cluster_Dtp2)) %>% 
    compute_plot_beta(ps_up = .,
                      beta = beta,
                      color_group = "Sample",
                      shape_group = "Sample",
                      alpha = NULL,
                      col_pal = sample_pal,
                      fill_pal = sample_pal,
                      path_group = "interaction(Subject)",
                      facet_formula = "Sample ~ .",
                      axis1 = 1,
                      axis2 = 2,
                      seed = 123,
                      permanova_terms = c("cluster_Dtp2"),
                      metadata_dist_boxplot = NULL ,
                      strata = "none") -> Time_out_saliva_clust[[tp]]
}

```

```{r}
Time_out_saliva_clust %>% 
  plyr::ldply(., function(x) x$perm) %>%
  rename(Time = '.id') %>% 
  DT::datatable()
```

Plaque

```{r,return = FALSE, message=FALSE, warning=FALSE}
tmp = ps_up %>% 
  subset_samples(Sample == "Plaque")
group = "Time"
Time_out_plaque_clust=NULL
# Time_out <- vector("list", length(tmp %>%
#                                get_variable(group) %>%
#                                levels()))
# names(Time_out) <- tmp %>%
#   get_variable(group) %>% levels()

for(tp in tmp %>%
    get_variable(group) %>%
    unique()){
  
  prune_samples(get_variable(tmp, group) == tp,
                tmp) %>%
    subset_taxa(Class != "UNCLASSIFIED") %>% 
    transform_sample_counts(function(x) x/sum(x) * 1) %>% 
    microViz::ps_mutate(cluster_Dtp2 = as.factor(cluster_Dtp2)) %>% 
    compute_plot_beta(ps_up = .,
                      beta = beta,
                      color_group = "Sample",
                      shape_group = "Sample",
                      alpha = NULL,
                      col_pal = sample_pal,
                      fill_pal = sample_pal,
                      path_group = "interaction(Subject)",
                      facet_formula = "Sample ~ .",
                      axis1 = 1,
                      axis2 = 2,
                      seed = 123,
                      permanova_terms = c("cluster_Dtp2"),
                      metadata_dist_boxplot = NULL ,
                      strata = "none") -> Time_out_plaque_clust[[tp]]
}

```

```{r}
Time_out_plaque_clust %>% 
  plyr::ldply(., function(x) x$perm) %>%
  rename(Time = '.id') %>% 
  DT::datatable()
```

```{r}
save(Time_out_plaque_clust, Time_out_saliva_clust, Time_out_clust, Time_out, plaque, saliva, d_box, d_box_leg, 
     ps_up,
     sample_pal,
     time_pal, sub_pal, sex_pal, sample_pal, period_pal,
     file = here::here("../../data/processed_data/metaphlan/03_data.Rdata"))
```


```{r}
sessionInfo()
```
